Performance Models for Data Transfers: A Case Study with Molecular Chemistry Kernels

### 2019-04-22 ICPP-19/communications.tex
Proofread

	With the advent of multicore, and the use of accelerators, it is notoriously cumbersome to exploit the full capability of a machine. Indeed, there are several challenges that come into picture. First, every architecture provides its own efficacy and interface. Therefore, a steep learning curve is required for programmers to take good utilization of all resources. Second, scheduling is a well known NP-Complete optimization problem, and hybrid and distributed resources make this problem harder (we refer ~\cite{webpagescheduling} for a survey on the complexity of scheduling problems and ~\cite{bleuse2015scheduling} for a recent survey in the case of hybrid nodes). Third, due to shared buses and parallel resources, it is challenging to obtain a precise model based on prediction of computation and communication times. Fourth, the number of architectures has increased drastically in recent years, therefore it is almost impossible to develop hand tuned optimized code for all these architectures. All these observations led to the development of different task based runtime systems. Among several runtimes, we may cite QUARK~\cite{YarKhan:2011:Quark:Manual} and PaRSEC~\cite{parsec} from from ICL, Univ. of Tennessee Knoxville (USA), StarPU~\cite{starpu} form Inria Bordeaux (France), Legion~\cite{legion12} from Stanford Univ. (USA), StarSs~\cite{ompss} from Barcelona Supercomputing Center (Spain), KAAPI~\cite{kaapi} from Inria Grenoble (France). All these runtime systems allow programmers to express their algorithms at the abstract level in the form of direct acyclic graphs (DAG), where vertices represent computations and edges represent dependencies among them. Sometimes some static information is also provided along with the DAG, such as distance to exit (last) node as a priority or affinity of computation towards resources. The runtime is then responsible for managing scheduling of computations and communications, data transfers among different memories, computation-communication overlap, and load balance.

@@

	Certain applications such as dense linear algebra kernels have regular structure. Therefore, it is possible to associate priorities to computations, based on the task graph structure, and to use them at runtime to make the execution efficient. In irregular applications, programmers do not know the precise structure of the task graphs in advance: tasks are added recursively based on certain sentinel constraints. For such applications, the runtime system sees a set of independent tasks and schedules them on processing units of different memory spaces. It is extremely important for runtimes to decide the order of data transfers for these scheduled computations so as to maximize the overlap between computations and communications. This is the main topic of this paper. We prove that the order of communications on two memory nodes with the objective of minimizing the makespan is a NP-Complete problem if the memory of the target node is limited. Our proof is inspired from work by~\cite{Papadimitriou:1980:FSL:322203.322213}, which applies a similar technique for 2-machine flowshop problem with bounded capacity. The main difference between both approaches is that they consider all tasks have the same occupation on the second machine and the memory occupation starts when the processing finishes on the first machine. On the contrary, our approach is designed for tasks appearing in scientific workloads whose memory requirements are highly irregular and we consider that memory is acquired before starting the data transfer on the communication resource. We propose different runtime strategies in order to maximize the overlap of computations and communications. We evaluate our strategies on the context of a cluster of homogeneous nodes. However, our approach is generic and easily adaptable to any system which operates on different memory spaces. Here are the important contributions of this paper:

	\renewcommand{\labelitemi}{$\bullet$}
	\begin{itemize}
		\item NP-Completeness proof for the general data-transfer problem 
		\item Proposed different scheduling strategies with the objective to minimize the makespan
		%%		\item An optimal strategy and its proof, when there is not any memory capacity constraint on the target machine
		\item Linear programming formulation of the problem
		\item Numerous experiments to assess the effectiveness of our strategies on molecular chemistry kernels 
	\end{itemize}

@@

	The problem of scheduling tasks has been highly studied in the literature and many formulations are known to be NP-Complete~\cite{GareyJohnson}. Many static and dynamic strategies have been proposed and analyzed for scheduling tasks on heterogeneous resources~\cite{heft-Topcuoglu,ipdps16starpu}. There is also a number of studies in the direction of task scheduling with the emphasis on improving locality and minimizing the communication cost~\cite{starpu,heft-Topcuoglu}. Stanisic et. al~\cite{luka-dmdar} proposed a heuristic to schedule tasks on a computational resource where most of its data is available. A similar approach has been adopted by Agullo et. al for the scheduling of sparse linear algebra kernels~\cite{agullo_fmm}. Predari et. al proposed heuristics to partition the task graph across a number of processors such that inter-processor communication can be minimized~\cite{predari:tel-01518956}.

	The problem considered in this paper also can be viewed as a flow shop problem: the communication link can be seen as a processing resource, and each task needs to first be handled by the communication link and then by the computational resource. Communication and computation times of a task can thus be considered as processing times on different machines. Johnson has provided scheduling strategies for 2 and 3-machine flow shop problems with infinite memory capacity~\cite{johnson}. 2-machine flow shop problem with finite buffer has been proven NP-Complete by Papadimitriou et. al~\cite{Papadimitriou:1980:FSL:322203.322213}, in which a constraint is imposed on the number of tasks that can await execution on the second machine.

@@

		In this paper, we consider the problem of deciding the order of data transfers between two memory nodes such that overlap of communications and computations is maximized. With Exascale computing, applications face bottlenecks due to communications. Hence, it is extremely important to achieve the maximum  communication-computation overlap in order to exploit the full potential of the system. We show that the problem of determining the optimal order of data transfers is NP complete. We propose several data transfer heuristics and evaluate them on two molecular chemistry kernels, HF and CCSD. Our results show that some of our heuristics achieve significant overlap and their makespans are very close to the lower bound. We plan to evaluate our strategies on different applications coming from multiple domains. We also plan to study the behavior of our strategies in the context of overlapping CPU-GPU communications with computations. A runtime system aiming to expose different heuristics to maximize the communication-computation overlap at the developer level and automatically select the best one is currently underway.

>>>>

	With the advent of multicore, and the use of accelerators, it is notoriously cumbersome to exploit the full capability of a machine. Indeed, there are several challenges that come into picture. First, every architecture provides its own efficacy and interface. Therefore, a steep learning curve is required for programmers to take good utilization of all resources. Second, scheduling is a well known NP-complete optimization problem, and hybrid and distributed resources make this problem harder (we refer ~\cite{webpagescheduling} for a survey on the complexity of scheduling problems and ~\cite{bleuse2015scheduling} for a recent survey in the case of hybrid nodes). Third, due to shared buses and parallel resources, it is challenging to obtain a precise model based on prediction of computation and communication times. Fourth, the number of architectures has increased drastically in recent years, therefore it is almost impossible to develop hand tuned optimized code for all these architectures. All these observations led to the development of different task based runtime systems. Among several runtimes, we may cite QUARK~\cite{YarKhan:2011:Quark:Manual} and PaRSEC~\cite{parsec} from from ICL, Univ. of Tennessee Knoxville (USA), StarPU~\cite{starpu} form Inria Bordeaux (France), Legion~\cite{legion12} from Stanford Univ. (USA), StarSs~\cite{ompss} from Barcelona Supercomputing Center (Spain), KAAPI~\cite{kaapi} from Inria Grenoble (France). All these runtime systems allow programmers to express their algorithms at the abstract level in the form of direct acyclic graphs (DAG), where vertices represent computations and edges represent dependencies among them. Sometimes some static information is also provided along with the DAG, such as distance to exit (last) node as a priority or affinity of computation towards resources. The runtime is then responsible for managing scheduling of computations and communications, data transfers among different memories, computation-communication overlap, and load balance.

@@

	Certain applications such as dense linear algebra kernels have regular structure. Therefore, it is possible to associate priorities to computations, based on the task graph structure, and to use them at runtime to make the execution efficient. In irregular applications, programmers do not know the precise structure of the task graphs in advance: tasks are added recursively based on certain sentinel constraints. For such applications, the runtime system sees a set of independent tasks and schedules them on processing units of different memory spaces. It is extremely important for runtimes to decide the order of data transfers for these scheduled computations so as to maximize the overlap between computations and communications. This is the main topic of this paper. We prove that the order of communications on two memory nodes with the objective of minimizing the makespan is a NP-complete problem if the memory of the target node is limited. Our proof is inspired from work by~\cite{Papadimitriou:1980:FSL:322203.322213}, which applies a similar technique for 2-machine flowshop problem with bounded capacity. The main difference between both approaches is that they consider all tasks have the same occupation on the second machine and the memory occupation starts when the processing finishes on the first machine. On the contrary, our approach is designed for tasks appearing in scientific workloads whose memory requirements are highly irregular and we consider that memory is acquired before starting the data transfer on the communication resource. We propose different runtime strategies in order to maximize the overlap of computations and communications. We evaluate our strategies on the context of a cluster of homogeneous nodes. However, our approach is generic and easily adaptable to any system which operates on different memory spaces. Here are the important contributions of this paper:

	\renewcommand{\labelitemi}{$\bullet$}
	\begin{itemize}
		\item NP-completeness proof for the general data-transfer problem 
		\item Proposed different scheduling strategies with the objective to minimize the makespan
		%%		\item An optimal strategy and its proof, when there is not any memory capacity constraint on the target machine
		\item Linear programming formulation of the problem
		\item Numerous experiments to assess the effectiveness of our strategies on molecular chemistry kernels 
	\end{itemize}

@@

	The problem of scheduling tasks has been highly studied in the literature and many formulations are known to be NP-complete~\cite{GareyJohnson}. Many static and dynamic strategies have been proposed and analyzed for scheduling tasks on heterogeneous resources~\cite{heft-Topcuoglu,ipdps16starpu}. There is also a number of studies in the direction of task scheduling with the emphasis on improving locality and minimizing the communication cost~\cite{starpu,heft-Topcuoglu}. Stanisic et. al~\cite{luka-dmdar} proposed a heuristic to schedule tasks on a computational resource where most of its data is available. A similar approach has been adopted by Agullo et. al for the scheduling of sparse linear algebra kernels~\cite{agullo_fmm}. Predari et. al proposed heuristics to partition the task graph across a number of processors such that inter-processor communication can be minimized~\cite{predari:tel-01518956}.

	The problem considered in this paper also can be viewed as a flow shop problem: the communication link can be seen as a processing resource, and each task needs to first be handled by the communication link and then by the computational resource. Communication and computation times of a task can thus be considered as processing times on different machines. Johnson has provided scheduling strategies for 2 and 3-machine flow shop problems with infinite memory capacity~\cite{johnson}. 2-machine flow shop problem with finite buffer has been proven NP-complete by Papadimitriou et. al~\cite{Papadimitriou:1980:FSL:322203.322213}, in which a constraint is imposed on the number of tasks that can await execution on the second machine.

@@

		In this paper, we consider the problem of deciding the order of data transfers between two memory nodes such that overlap of communications and computations is maximized. With Exascale computing, applications face bottlenecks due to communications. Hence, it is extremely important to achieve the maximum  communication-computation overlap in order to exploit the full potential of the system. We show that the problem of determining the optimal order of data transfers is NP-complete. We propose several data transfer heuristics and evaluate them on two molecular chemistry kernels, HF and CCSD. Our results show that some of our heuristics achieve significant overlap and their makespans are very close to the lower bound. We plan to evaluate our strategies on different applications coming from multiple domains. We also plan to study the behavior of our strategies in the context of overlapping CPU-GPU communications with computations. A runtime system aiming to expose different heuristics to maximize the communication-computation overlap at the developer level and automatically select the best one is currently underway.
